---
title: "Machine Learning with Activity Monitoring Data"
author: "JMFlin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

As described in the website, six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. 
 
```{r, message=FALSE}
library(ggplot2)
library(caret)
library(reshape2)
```

```{r}
training <- read.csv("training.csv", na.string = c("NA","Na", "NaN", "", " ", "#DIV/0!"))
validation <- read.csv("validation.csv", na.string = c("NA","Na", "NaN", "", " ", "#DIV/0!"))
```
 
Remove those columns that have 70% of their values missing. Check for zero and near zero variance predictors and remove highly correlated predictors.
 
```{r}
train <- training[, -which(colMeans(is.na(training)) > 0.7)]

train <- train[, !grepl("^X|timestamp|window", names(train))]

nearZeroVar(train, saveMetrics = TRUE)

it <- findCorrelation(cor(train[,2:(ncol(train)-1)]), .70) 
it <- it+1
train <- train[, -it]
```
 
Do some visual checking of the data. We can see from gyros_forearm_x that there is an outlier in the data. This outlier is in the row numbered 5373. I will remove it from the data.
 
```{r}
temp <- train
temp <- temp[,2:17]
temp <- cbind(temp, classe = train[,"classe"])
longtemp <- melt(temp, measure.vars = (1:ncol(temp)-1))
ggplot(longtemp, aes(x=classe, y=value))+
  xlab("")+
  ylab("")+
  stat_boxplot(geom ='errorbar')+ 
  geom_boxplot() +
  theme(axis.line = element_line(), axis.text=element_text(color='black'), axis.title = element_text(colour = 'black'), legend.text=element_text(), legend.title=element_text(), legend.key = element_rect(colour = "black"))+
  facet_wrap(~ variable, ncol = 4, scales = "free")

temp <- train
temp <- temp[,18:ncol(train)]
longtemp <- melt(temp, measure.vars = (1:ncol(temp)-1))
ggplot(longtemp, aes(x=classe, y=value))+
  xlab("")+
  ylab("")+
  stat_boxplot(geom ='errorbar')+ 
  geom_boxplot() +
  theme(axis.line = element_line(), axis.text=element_text(color='black'), axis.title = element_text(colour = 'black'), legend.text=element_text(), legend.title=element_text(), legend.key = element_rect(colour = "black"))+
  facet_wrap(~ variable, ncol = 4, scales = "free")

row.names(train[train$gyros_forearm_x < -10,])
train <- train[-as.numeric(row.names(train[train$gyros_forearm_x < -10,])),]
```
 


First we split the data into training and testing sets. Then we set the seed so that the analysis can be reproduced exactly. Then we use the train function to set up 5-fold cross validation and train a random forest algorithm on the training data using 250 trees.
 
```{r, message=FALSE}
validation <- validation[,names(validation) %in% names(train)]

index <- createDataPartition(y=train$classe, p=0.7, list=FALSE)
train_split <- train[index, ]
train_test_split <- train[-index, ]

ctrl <- trainControl(method = "cv",
                     number = 5)

set.seed(1234)

rf_mod <- train(classe ~ ., 
                data = train_split,
                method = "rf",
                ntree = 250,
                trControl = ctrl)

rfClasses <- predict(rf_mod, train_test_split)
```

The accuracy of the model is about 0.99. We can be fairly confident that we will predict the class of all 20 observations with perfectly.
 
```{r}
confusionMatrix(data = rfClasses, train_test_split$classe)
```
 
The predicted classes of these 20 unknown observations are shown here:
 
```{r}
result <- predict(rf_mod, validation)
result
```